shorts: An R Package for Modeling Short Sprints

### 2021-11-20 shorts-paper.Rmd
Paper prepared for submission to International Journal of Strength and Conditioning after review comments

---
title: "shorts: An R Package for Modeling Short Sprints"
preprint: true
date: "`r Sys.Date()`"
author: 
  - name: Mladen Jovanović
    affiliation: 1
    corresponding: true
    email: coach.mladen.jovanovic@gmail.com
  - name: Jason D. Vescovi
    affiliation: 2
    corresponding: false
    email: vescovij@gmail.com
affiliation:
  - code: 1
    address: Faculty of Sport and Physical Education, University of Belgrade, Serbia
  - code: 2
    address: Faculty of Kinesiology and Physical Education, Graduate School of Exercise Science, Toronto, ON Canada
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the **shorts** package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the **shorts** package to help alleviate errors within the subsequent parameter outcomes.
header-includes: >
  \usepackage{lipsum}
  \usepackage[htt]{hyphenat}
  \usepackage{booktabs}
  \usepackage{microtype}
preamble: >
  \usepackage{amsmath}
citation_package: natbib
bibliography: [references.bib, packages.bib]
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    extra_dependencies: ["flafter"]
  rticles::peerj_article: default
link-citations: yes
editor_options:
  chunk_output_type: console
---

@@

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 300,
  out.width = "90%",
  fig.align = "center",
  fig.width = 5,
  fig.height = 5 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 65,
  size = "small"
)

@@

Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

@@

Short sprints have been modeled using the mono-exponential equation \@ref(eq:velocity-time) originally proposed by @doi:10.1098/rspb.1927.0035, and more recently popularized by @clarkNFLCombine40Yard2017, and @samozinoSimpleMethodMeasuring2016. Equation \@ref(eq:velocity-time) represents function for instantaneous horizontal velocity $v$ given the time $t$ and two model parameters:

@@

By derivating equation \@ref(eq:velocity-time), we can get equation for horizontal acceleration \@ref(eq:acceleration-time).

@@

By integrating equation \@ref(eq:velocity-time), we can get equation for distance covered \@ref(eq:distance-time).

@@

```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in **shorts** package using non-linear least squares regression implemented in the `nls()` function [@batesNonlinearModels1992; @batesNonlinearRegressionAnalysis2007] in the **base R** [@R-base] and `nlme()` function in the **nlme** package [@R-nlme] for the mixed-effect models.

@@

$W$ in equation \@ref(eq:time-distance) represents Lambert's W function [@R-LambertW]. Researchers often incorrectly use the equation \@ref(eq:distance-time) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothSpreadsheetSprintAcceleration2020], in which the time is the predictor and distance is the outcome variable, instead of statistically correct equation \@ref(eq:time-distance) [@motulskyIntuitiveBiostatisticsNonmathematical2018, p. 341].

@@

Both split and radar gun models allow the use of *weighted* non-linear regression. For example, we can give more weight to shorter distance or faster velocities. Weighted non-linear regression is performed by setting `weights` parameter:

@@

```{r}
require(knitr)
require(kableExtra)

@@

kable(jack_parameters, digits = 2, booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```

@@

kable(adding_zero_issue, digits = 2, booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```

@@

It is important to note that using `predict_XXX()` family of functions with distance correction model and using estimated distance correction parameter will yield wrong predictions on the original scale (but not on the *theoretical* scale in which force production starts at d=0m and t=0s). This is because `predict_XXX()` family of functions utilize equation \@ref(eq:time-distance-correction) and its derivates and integrals, which works for the uncorrected model (time and distance correction parameters are equal to zero), time correction models (distance correction parameter is equal to zero), and time and distance corrections model. You can notice that equations \@ref(eq:time-correction) and \@ref(eq:time-distance-correction) are equivalent to the \@ref(eq:time-distance) equation when time and distance correction parameters are equal to zero. Unfortunately, this doesn't work for the distance correction model when estimated distance correction parameter is used since model definition outlined in the equation \@ref(eq:distance-correction) is different that the equation \@ref(eq:time-distance-correction) when estimated distance correction parameter is different than zero. The solution is to use generic S3 `predict()` function, but this only predicts split times at split distances.

@@

# To predict at "theoretical" distance scale
#  (where force application starts at d=0m)
predict_time_at_distance(
  distance = split_times$distance, 
  MSS = jack_profile_time_estimated$parameters$MSS,
  TAU = jack_profile_time_estimated$parameters$TAU)

@@

To estimate parameter stability, model over-fitting, and performance on the unseen data, **shorts** model function comes with implemented *leave-one-out cross validation* (LOOCV) [@jamesIntroductionStatisticalLearning2017; @jovanovicBmbstatsBootstrapMagnitudebased2020; @kuhnAppliedPredictiveModeling2018]. LOOCV involves a simple, yet powerful procedure, of removing each observation, rebuilding the model, and making predictions for that removed observation. This process is repeated for each observation in the model dataset. LOOCV allows one to check estimated parameters stability, and model performance on the unseen data.

@@

Let's utilize demonstrated functionalities of the **shorts** package using real-world data. The first dataset comes from Usain Bolt's performance from IAAF World Championship held in London, 2017, and the second dataset involve Jason Vescovi's sample data for 52 female soccer and field hockey athletes which comes with the **shorts** package (see `?vescovi`).

@@

kable(bolt_model, digits = 2, booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```

@@

The following models were used: (1) no corrections model, (2) fixed time correction model (using 0.3s heuristic rule of thumb), (3) time correction model with time correction as fixed effect, (4) time correction model with time correction as random effect, (5) distance correction model with distance correction as fixed effect, (6) distance correction model with distance correction as random effect, (7) time and distance correction model, (8) time and distance correction model with distance correction as fixed effect (and time correction as random effect), and (9) time and distance correction model with distance and time correction as random effects (please refer to the [Supplemental material] for the R code).

>>>>

---
title: "shorts: An R Package for Modeling Short Sprints"
date: "`r Sys.Date()`"
author: 
  - Mladen Jovanović^[Faculty of Sport and Physical Education, University of Belgrade, Serbia, coach.mladen.jovanovic@gmail.com]
  - Jason D. Vescovi^[Faculty of Kinesiology and Physical Education, Graduate School of Exercise Science, Toronto, ON Canada, vescovij@gmail.com]
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the **shorts** package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the **shorts** package to help alleviate errors within the subsequent parameter outcomes.
citation_package: natbib
bibliography: [references.bib, packages.bib]
output:
  bookdown::word_document2:
    number_sections: FALSE
link-citations: yes
editor_options:
  chunk_output_type: console
---

@@

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = FALSE,
  fig.retina = 0.8, # figures are either vectors or 300 dpi diagrams
  dpi = 600,
  out.width = "90%",
  fig.align = "center",
  fig.width = 5,
  fig.height = 5 * 0.618, # 1 / phi
  fig.show = "hold",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  width = 65,
  size = "small"
)

@@

Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the athlete and their sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

@@

Short sprints have been modeled using the mono-exponential equation \@ref(eq:velocity-time) originally proposed by @doi:10.1098/rspb.1927.0035, and more recently popularized by @clarkNFLCombine40Yard2017, and @samozinoSimpleMethodMeasuring2016. Equation \@ref(eq:velocity-time) represents the function for instantaneous horizontal velocity $v$ given the time $t$ and two model parameters:

@@

By derivating equation \@ref(eq:velocity-time), we can get the equation for horizontal acceleration \@ref(eq:acceleration-time).

@@

By integrating equation \@ref(eq:velocity-time), we can get the equation for distance covered \@ref(eq:distance-time).

@@

```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  xlim(c(0, 12)) +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), or (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in **shorts** package using non-linear least squares regression implemented in the `nls()` function [@batesNonlinearModels1992; @batesNonlinearRegressionAnalysis2007] in the **base R** [@R-base] and `nlme()` function in the **nlme** package [@R-nlme] for the mixed-effect models.

@@

$W$ in equation \@ref(eq:time-distance) represents Lambert's W function [@R-LambertW]. Researchers often incorrectly use the equation \@ref(eq:distance-time) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothSpreadsheetSprintAcceleration2020], in which the time is the predictor and distance is the outcome variable, instead of statistically correct equation \@ref(eq:time-distance). This practice should be avoided because switching of the predictor and outcome variables in the regression model might produce biased estimated parameters [@motulskyIntuitiveBiostatisticsNonmathematical2018, p. 341]. This might not necessarily be the case with the short sprints profiling, but it is nevertheless a bad statistical practice.

@@

Both split and radar gun models allow the use of *weighted* non-linear regression. Weighting in regression is utilized when the observations have unequal error variance [@gelmanRegressionOtherStories2020], which can happen due to instrumental error (e.g., multiplicative error instead of additive error) or due to biological variability (e.g., higher split time variance on shorter distances versus longer distances). In this case, observations with higher error variance get lower weight when fitting the model [@gelmanRegressionOtherStories2020]. According to @gelmanRegressionOtherStories2020, unequal variances are not typically a major issue for the goal of estimating regression parameters, but they can become more important when making predictions about individual observations. Further exploring the topic of unequal error variances is beyond the scope of this paper, but the **shorts** package provides weighting feature for the potential research of this topic in the future.

Weighted non-linear regression is performed by setting `weights` parameter. For example, we can give more weight to shorter distance or faster velocities:

@@

```{r}
require(knitr)

@@

kable(jack_parameters, digits = 2, booktabs = TRUE)
```

@@

kable(adding_zero_issue, digits = 2, booktabs = TRUE)
```

@@

It is important to note that using `predict_XXX()` family of functions with distance correction model and using estimated distance correction parameter will yield wrong predictions on the original scale (but not on the *theoretical* scale in which force production starts at d=0m and t=0s). This is because `predict_XXX()` family of functions utilize equation \@ref(eq:time-distance-correction) and its derivates and integrals, which works for the uncorrected model (time and distance correction parameters are equal to zero), time correction models (distance correction parameter is equal to zero), and time and distance corrections model. You can notice that equations \@ref(eq:time-correction) and \@ref(eq:time-distance-correction) are equivalent to the \@ref(eq:time-distance) equation when time and distance correction parameters are equal to zero. Unfortunately, this doesn't work for the distance correction model when the estimated distance correction parameter is used since model definition outlined in the equation \@ref(eq:distance-correction) is different that the equation \@ref(eq:time-distance-correction) when the estimated distance correction parameter is different than zero. The solution is to use generic S3 `predict()` function, but this only predicts split times at split distances.

@@

# To predict at "theoretical" distance scale (where force application starts at d=0m)
predict_time_at_distance(
  distance = split_times$distance, 
  MSS = jack_profile_time_estimated$parameters$MSS,
  TAU = jack_profile_time_estimated$parameters$TAU)

@@

To estimate parameter stability, model over-fitting, and performance on unseen data, **shorts** model function comes with implemented *leave-one-out cross validation* (LOOCV) [@jamesIntroductionStatisticalLearning2017; @jovanovicBmbstatsBootstrapMagnitudebased2020; @kuhnAppliedPredictiveModeling2018]. LOOCV involves a simple, yet powerful procedure, of removing each observation, rebuilding the model, and making predictions for that removed observation. This process is repeated for each observation in the model dataset. LOOCV allows one to check estimated parameters stability, and model performance on unseen data.

@@

Let's utilize demonstrated functionalities of the **shorts** package using real-world data. The first dataset comes from Usain Bolt's performance from IAAF World Championship held in London, 2017, and the second dataset involve Jason Vescovi's sample data for 52 female soccer and field hockey athletes which comes with the **shorts** package (see the dataset documentation by executing `?vescovi` command in the R console).

@@

kable(bolt_model, digits = 2, booktabs = TRUE)
```

@@

The following models were used: (1) no corrections model, (2) fixed time correction model (using +0.3s heuristic rule of thumb), (3) time correction model with time correction as fixed effect, (4) time correction model with time correction as random effect, (5) distance correction model with distance correction as fixed effect, (6) distance correction model with distance correction as random effect, (7) time and distance correction model, (8) time and distance correction model with distance correction as fixed effect (and time correction as random effect), and (9) time and distance correction model with distance and time correction as random effects (please refer to the [Supplemental material] for the R code).
